Male-biased sexual selection, but not sexual dichromatism, predicts speciation in birds

### 2021-01-18 title_page.tex
add tex

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath} \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling} \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large} \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[L]{\textsl{\leftmark}} \rhead[]{Sexual selection predicts speciation} \usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno} \linenumbers \usepackage{setspace}\onehalfspacing \usepackage[font={small}]{caption} \usepackage{epigraph} \setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt} \setlength\beforeepigraphskip{1\baselineskip} \setlength\afterepigraphskip{2\baselineskip} \renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape} \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{} \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{} \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} \usepackage[hang, flushmargin]{footmisc} \usepackage{hyperref}\hypersetup{colorlinks=true, hyperfootnotes=false, linkcolor=blue, filecolor=blue, urlcolor=blue} \usepackage{xcolor} \usepackage{pdflscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}} \raggedbottom

\begin{document}
\maketitle
\begin{abstract}
Sexual selection is thought to shape phylogenetic diversity by affecting
speciation or extinction rates. However, the net effect of sexual
selection on diversification is hard to predict, because many of the
hypothesised effects on speciation or extinction have opposing signs and
uncertain magnitudes. Theoretical work also suggests that the net effect
of sexual selection on diversification should depend strongly on
ecological factors, though this prediction has seldom been tested. Here,
we test whether variation in sexual selection can predict speciation and
extinction rates across passerine birds (up to 5,812 species, covering
most genera) and whether this relationship is mediated by environmental
factors. Male-biased sexual selection, and specifically sexual size
dimorphism, predicted two of the three measures of speciation rates that
we examined. The link we observed between sexual selection and
speciation was independent of environmental variability, though species
with smaller ranges had higher speciation rates. There was no
association between any proxies of sexual selection and extinction rate.
Our findings support the view that male-biased sexual selection, as
measured by frequent predictors of male-male competition, has shaped
diversification in the largest radiation of birds.
\end{abstract}

@@

\newpage
\phantomsection

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Sexual selection is a ubiquitous evolutionary process whose effect on
phylogenetic diversification is much debated (Lande 1981, 1982;
West-Eberhard 1983; Seddon et al. 2008; Cooney et al. 2018; Tsuji and
Fukami 2020). Sexual selection can promote speciation because it
operates on traits that can create reproductive isolation when they
diverge between lineages, such as signals and preferences involved in
mate selection (Lande 1981, 1982; Safran et al. 2013), sperm-egg
interactions (Swanson and Vacquier 1998), or genital morphology (Sloan
and Simmons 2019). Sexual selection could also promote speciation or
prevent extinction by purging deleterious mutations (Whitlock and
Agrawal 2009), fixing beneficial ones (Whitlock 2000), and accelerating
adaptation in different environments (Lorch et al. 2003; Candolin and
Heuschele 2008; Cally et al. 2019). Conversely, sexual selection might
hinder speciation or make extinction more likely by favouring traits
that improve mating success but reduce population fitness (Kokko and
Jennions 2008; Rankin et al. 2011; Holman and Kokko 2013; Fromhage and
Jennions 2016). For example, species with costly sexual signals may be
less resilient to environmental change (Kokko and Brooks 2003).
Extinction risk may also be exacerbated by sexual selection causing
maladaptation (`gender load') in female traits that are genetically
correlated with sexually-selected male traits (Pischedda and Chippindale
2006; Bonduriansky and Chenoweth 2009; Harano et al. 2010; Pennell and
Morrow 2013; Berger et al. 2014).

Although numerous studies have examined the relationship between sexual
selection and speciation or extinction rates (Barraclough et al. 1995;
Morrow et al. 2003; Seddon et al. 2008, 2013; Kraaijeveld et al. 2011;
Huang and Rabosky 2014), the availability of more complete phenotypic,
ecological and phylogenetic data (Jetz et al. 2012), together with
significant advances in phylogenetic methods (Rabosky 2014; Harvey
Michael et al. 2017), present new opportunities to test whether and how
sexual selection drives diversification. Furthermore, the diversity of
outcomes and approaches in previous studies suggests that the
association between species diversity and sexual selection is far from
clear (reviewed in Tsuji and Fukami (2020)).

A possible reason for the above uncertainty regarding the relationship
between sexual selection and diversification is that this relationship
may strongly depend on the environment. Theoretical work predicts that
sexual selection should have a more positive effect on adaptation and
population fitness in variable environments relative to stable ones
(Long et al. 2012; Connallon and Hall 2016). In stable environments,
consistent selection depletes genetic variation at sexually concordant
loci (i.e.~loci where the same allele is fittest for both sexes). In
these environments, genetic variation remains disproportionately at
sexually antagonistic loci, leading to stronger gender load and reduced
net benefits of sexual selection (Connallon and Hall 2016). By contrast,
in spatially or temporally variable environments, sexual selection can
enhance local adaptation. For example, in Darwin's finches, divergent
beak morphology is an adaptation to local food availability that has
been maintained through assortative mating (Huber et al. 2007). Under
these circumstances we predict that the effect of sexual selection on
rates of divergence may depend on the variability of the species'
environment. Despite the potential interaction between sexual selection
and environmental variability in diversification, phylogenetic tests are
currently lacking.

Birds have been a popular focus of macroevolutionary studies of sexual
selection and diversification (Barraclough et al. 1995; Morrow et al.
2003; Seddon et al. 2008, 2013; Huang and Rabosky 2014) because their
biology and phylogenetic relationships are comparatively well-known. A
2011 meta-analysis covering 20 primary studies of birds and other taxa
found a small but significant positive association between sexual
selection and speciation, with the average effect size in birds stronger
than in mammals but weaker than in insects or fish (Kraaijeveld et al.
2011). However, there was large variation in effect size estimates
across the 20 studies, likely reflecting differences in methodology,
such as metrics used to characterise speciation and sexual selection, in
addition to true biological differences. More recently, Huang and
Rabosky (2014) found no association between sexual dichromatism and
speciation (\emph{n} = 918 species) in a study using spectrophotometric
measurements of museum specimens (Armenta et al. 2008) and tip-rate
estimates from a molecular-only phylogeny (Jetz et al. 2012). Similarly,
Cooney et al. (2017) found no effect of sexual dichromatism on
diversification across 1,306 pairs of species, using dichromatism scores
provided by human observers. More recently, social polygyny (a proxy for
sexual selection) was found to have a positive association with
speciation rate across 954 species of birds (Iglesias-Carrasco et al.
2019). We summarize the major findings from previous studies testing the
association between sexual selection and speciation in birds and other
taxa since Kraaijeveld et al. (2011) meta-analysis (Table 1).

Here, we investigate the association between sexual selection and
diversification in birds while building upon previous approaches in
multiple ways. We use two measures of the strength of sexual selection:
sexual dichromatism (Dale et al. 2015), as well as an index of
male-biased sexual selection (Dale et al. 2015), which captures
(co)variation in sexual size dimorphism, social polygyny and paternal
care. We use these two measures because sexual dichromatism does not
always signal the presence of strong sexual selection and \emph{vice
versa} (Dale et al. 2015). For example, male and female dunnocks
(\emph{Prunella modularis}) are similarly coloured yet sexual selection
appears to be strong (Davies and Houston 1986). Furthermore, a recent
comparative study found a negative relationship between dichromatism and
another sexually-selected trait (song) across species, suggesting that a
multi-trait focus would improve estimates of sexual selection intensity
(Cooney et al. 2018). Additionally, our analysis includes multiple
ecological and environmental variables, allowing us to control for
potential confounds, to identify environmental factors, including
spatial and temporal environmental variability, interact with sexual
selection as theory predicts (Connallon and Hall 2016).

We use multiple approaches for quantifying speciation and extinction
rates at the tips of phylogenetic trees, including Bayesian Analysis of
Macroevolutionary Mixtures (BAMM; Beaulieu and O'Meara 2015; Rabosky
2016; Moore et al. 2016; Rabosky et al. 2017), as well as older but
reliable tip-rate statistics, namely diversification rate
(\(\lambda_{DR}\)) and node density (\(\lambda_{ND}\)) (Jetz et al.
2012). Our results show that (\emph{i}) a composite measure of sexual
selection, but not sexual dichromatism, significantly predicts
speciation rates, (\emph{ii}) the significant association between the
composite measure of sexual selection and speciation rate is largely
driven by sexual size dimorphism, (\emph{iii}) species with smaller
ranges have higher speciation rates and (\emph{iv}) there is no evidence
that environmental variables or their interaction with sexual selection
have an impact on diversification rates. Therefore, we provide evidence
at a very large scale that sexual selection can have positive effects on
diversification in the largest radiation of birds. Furthermore, we
suggest that the use of sexual dichromatism as the sole proxy for sexual
selection should be reconsidered, since it appears to be inconsistently
associated with the operation of sexual selection.

\begin{table}[H]

\caption{\label{tab:lit}Previous studies testing the association between sexual selection and speciation}
\centering
\resizebox{\linewidth}{!}{
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}{ll>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{3em}>{\raggedright\arraybackslash}p{27em}}
\toprule
Study & Taxa studied & Proxy for sexual selection & Support? & Outcome\\
\midrule
 &  & Plumage dichromatism & Yes & Across all birds, evidence in 4/6 studies\\
\cmidrule{3-5}
 &  & Mating system & Yes & Across all birds, evidence in 4/4 studies\\
\cmidrule{3-5}
\multirow{-3}{*}{\raggedright\arraybackslash Kraaijeveld et al. (2011)} & \multirow{-3}{*}{\raggedright\arraybackslash Meta-analysis across all animals} & Size dimorphism & Mixed & Across all birds, evidence in 1/2 studies\\
\cmidrule{1-5}
Maia et al. (2013) & Starlings (Sturnidae), 113 species & Ornamental innovations & Yes & Lineages with derived melanosomes (an ornamental innovation) diversify faster\\
\cmidrule{1-5}
Huang \& Rabosky (2014) & Across birds, \textasciitilde{}1000 species & Plumage dichromatism & No & No association between different measures of dichromatism and diversification\\
\cmidrule{1-5}
Gomes et al. (2016) & Estrildid finches, 134 species & Colour ornamentation & No & More ornamented lineages do not speciate more (but ornaments do evolve faster)\\
\cmidrule{1-5}
Cooney et al. (2017) & Across birds, 1306 pairs of species & Plumage dichromatism & No & Plumage dichromatism does not predict diversification rates, but might reduce the rate of fusion of lineages after secondary contact\\
\cmidrule{1-5}
Janicke et al. (2018) & Meta-analysis across all animals & Bateman gradient & Yes & Steepness of Bateman gradient in males predicts species richness\\
\cmidrule{1-5}
Mason et al. (2017) & Thraupids and Furnariids, 581 species & Vocal evolution & Yes & Bursts of speciation and song evolution are coincident\\
\cmidrule{1-5}
Iglesias-Carrasco et al. (2019) & Across birds, 954 species & Degree of polygyny & Yes & A higher degree of polygyny and rapid molecular evolution are linked with rate of diversification\\
\cmidrule{1-5}
Hosner et al. (2020) & Gallopheasants, 22 species & Sexual dimorphisn (range of traits) & No & No role of sexual selection in relation to diversification\\
\cmidrule{1-5}
Price-Waldman et al. (2020) & Thraupidae, 355 species & Plumage complexity & Yes & Elevated rates of plumage complexity evolution are associated with higher speciation rates\\
\cmidrule{1-5}
 &  & Size dimorphism & Yes & Sexual size dimorphism predicts two out of three measures of speciation rates\\
\cmidrule{3-5}
\multirow{-2}{*}{\raggedright\arraybackslash This study} & \multirow{-2}{*}{\raggedright\arraybackslash Across passerines, 5812 species} & Plumage dichromatism & No & There was no link between plumage dichromatism (measured from spectral info or RGB values) and any speciation rate\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item Studies were obtained by searching 'Web of Science' for articles published from 2011 for terms containing ‘speciation’, ‘diversification’ and ‘sexual selection’. We summarised all the studies we found relevant and comparable to our study.
\end{tablenotes}
\end{threeparttable}}
\end{table}

\hypertarget{materials-and-methods}{%
\section{Materials and methods}\label{materials-and-methods}}

We examined the effect of sexual selection on speciation and extinction
rate in 97\% of passerines (\emph{n} = 5,812 species; 58\% of all
birds). Specifically, we (\emph{i}) compiled datasets for sexual
dichromatism/selection strength and environmental variability,
(\emph{ii}) obtained estimates of speciation and extinction rates across
passerines, and (\emph{iii}) conducted phylogenetic generalized
least-squares (PGLS) regressions. Analyses are documented with
reproducible code in the
\href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information}.
\phantomsection

\hypertarget{compiling-data-for-sexual-selection-and-environmental-stress}{%
\subsection{Compiling data for sexual selection and environmental
stress}\label{compiling-data-for-sexual-selection-and-environmental-stress}}

\hypertarget{sexual-dichromatism}{%
\subsubsection{Sexual dichromatism}\label{sexual-dichromatism}}

We used a previously-published measure of sexual dichromatism for 5,983
species of passerines (Dale et al. 2015). Briefly, Dale et al. (Dale et
al. 2015) obtained sex-specific RGB (red-green-blue) values across six
body patches (nape, crown, forehead, throat, upper breast, and lower
breast) from \emph{Handbook of the Birds of the World} (Del Hoyo et al.
2011). The relative contribution of male and female RGB colour values
were averaged across body patches and provide `male-like' and
`female-like' plumage scores. Here we use the absolute difference
between male and female plumage scores as an estimate of sexual
dichromatism. Technically, this measures differences in the `degree of
male-ness' between males and females, rather than sex differences in
colour \emph{per se} (i.e.~dichromatism in the strict sense). For
example, the metric would fail to capture dichromatism when both the
male and female possess a single, but differently coloured `male-like'
patch. However, the metric is highly correlated with dichromatism
measured from spectral data (see below).

Additionally, we used another measure of dichromatism corresponding to
colour distance in avian colour space derived from spectral data
(Armenta et al. 2008). These measurements include variation in the
ultraviolet and bird-visible range and --- unlike the RGB measures ---
are sourced from museum specimens as opposed to illustrations. The
spectrophotometry data covers only 581 passerine species (10-fold fewer
than the RGB data), although there was a substantial correlation between
the two dichromatism measures (\(r = 0.79\);
\href{https://justincally.github.io/SexualSelection_Speciation/#subsetted_analysis_with_spectrophotometry_data}{Figure S10}).
\phantomsection

\hypertarget{male-biased-sexual-selection}{%
\subsubsection{Male-biased sexual
selection}\label{male-biased-sexual-selection}}

Sexual dichromatism is likely to be imperfectly correlated with
variation in the strength of sexual selection across taxa. For this
reason, we sourced an additional measure of sexual selection (Dale et
al. 2015), referred to here as the `index of male-biased sexual
selection'. This index is the first principal component from a
phylogenetic principal component analysis (PPCA) of three
characteristics possitively associated with sexual selection (sexual
size dimorphism, social polygyny and {[}lack of{]} paternal care). The
variables included in this index have all been positively linked to the
intensity of sexual selection, and are usually correlated (Björklund
1990; Owens and Hartley 1998; Dunn et al. 2001), which is why they were
combined into a single metric in previous studies (Dale et al. 2015).
This measure of male-biased sexual selection is available for only 2,465
species, and shows a moderate correlation with the RGB measure of sexual
dichromatism (\(r = 0.34\);
\href{https://justincally.github.io/SexualSelection_Speciation/#analysis_using_male-biased_measure_of_sexual_selection}{Figure S12}).
\phantomsection

\hypertarget{environmental-variables}{%
\subsubsection{Environmental variables}\label{environmental-variables}}

We obtained estimates of species range size using expert range maps
(BirdLife International and Handbook of the Birds of the World 2017).
The names of 1,230 species in the Birdlife database (Hoyo and Collar
2016) have been recently changed, so we manually matched these taxa with
the names used in the sexual dichromatism dataset (Hoyo and Collar
2016). For each species' range, we obtained estimates of climatic
conditions by extracting 1,000 random point samples of each bioclimatic
variable. We extracted 19 present-day bioclimatic variables
(representing a variety of biologically relevant annual trends in
temperature and precipitation) with 30-second (\textasciitilde{}1
km\textsuperscript{2}) spatial resolution (Fick and Hijmans 2017). From
the 1000 values of each bioclimatic variable, we obtained means and
standard deviations for each species. Using the same spatial sampling,
we extracted means and standard deviations of bioclimatic variables from
the paleoclimate during the last interglacial (LIG; 120,000 - 140,000
years ago) (Otto-Bliesner et al. 2006). To estimate variability in the
energy available to species, we obtained the mean and standard deviation
of net primary productivity (NPP) values between 2000 - 2015 across each
species distribution. Estimates of NPP had 30-second resolution and were
obtained through MODIS (Moderate Resolution Imaging Spectroradiometer)
primary production products stage 3 (MOD17A3) (Zhao et al. 2005). We
provide these data as a potentially useful data resource (see
\href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information}).
\phantomsection

\hypertarget{generating-biologically-relevant-predictors-for-environmental-stress}{%
\subsubsection{Generating biologically relevant predictors for
environmental
stress}\label{generating-biologically-relevant-predictors-for-environmental-stress}}

Given that stressful environments are expected to interact with sexual
selection and have a positive effect on adaptation (Cally et al. 2019),
we used the extracted environmental variables from each species range
size to create predictors of environmental variation/stress. We used
(\emph{i}) the average NPP in each species' range and (\emph{ii}) the
log-transformed range size as potentially informative predictors of
speciation rates. We also used three environmental predictors derived
from bioclimatic data. These predictors relate to seasonal climate
variation, spatial climate variation and long-term climate variation. To
obtain seasonal climate variation we used (\emph{iii}) mean values of
temperature seasonality (BIO4) for each range. (\emph{iv}) To estimate
levels of spatial environmental variation a species may endure, we used
the first principle component (PC1) from a PCA on standard deviations
from all bioclimatic variables, excluding temperature and precipitation
seasonality (BIO4 and BIO15). PC1 was heavily loaded towards bioclimatic
variables relating to temperature, thus PC1 largely reflects the
variation in temperature across a species' range
(\href{https://justincally.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Table S1}).
A taxon's range size often correlates with speciation and extinction
rates (Rosenzweig 1995; Castiglione et al. 2017), so we controlled for
the correlation between environmental spatial variation and species'
range sizes --- where larger ranges have larger variation in PC1 --- by
using the residuals of a fitted general additive model (GAM;
\href{https://justincally.github.io/SexualSelection_Speciation/#generating_biologically_relevent_predictors}{Figure S1})
as a predictor. To obtain long-term variation in climates for each
species range, we took (\emph{v}) the first principal component of the
absolute difference in the bioclimatic variables between the LIG and
current values. Similarly to spatial variation, the long-term climate
variation is primarily loaded to temperature differences between the LIG
and current climates
(\href{https://justincally.github.io/SexualSelection_Speciation/#long_term_climate_variability_(lig_anomaly)}{Table S2, Figure S2}).
The five predictors of environmental variability are not strongly
correlated
(\href{https://justincally.github.io/SexualSelection_Speciation/#correlations_between_environmental_predictors}{Figure S3}).
Details and \texttt{R} code to generate these predictors can be found
within the
\href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information}.
\phantomsection

\hypertarget{estimating-extinction-and-speciation}{%
\subsection{Estimating extinction and
speciation}\label{estimating-extinction-and-speciation}}

Phylogenetic information was obtained from www.birdtree.org (Jetz et al.
2012). We used a maximum clade credibility (MCC) tree from 2,500 samples
of the posterior distribution (\emph{n} = 5,965) as the main
phylogenetic hypothesis in our comparative analysis. Additionally, a
random draw of full trees (including species without genetic data) from
the posterior distribution of phylogenetic trees was used for
diversification analyses using tip-rate measures (1,000 trees) and BAMM
(100 trees) (Rabosky 2014). These trees had crown clades with a topology
that was heavily constrained on the basis of a previously published
study (``Hacket backbone''; Hackett et al. 2008) and were constructed
using a pure birth (Yule) model. We calculated three different tip-rate
metrics of speciation and one of extinction across all trees.

Diversification is the result of two processes, speciation and
extinction through time. To estimate speciation rates, we first obtained
two tip-rate metrics of speciation using statistics derived from the
properties of the nodes and branches along root-to-tip paths of the
phylogeny. Node density (ND) is a simple statistic calculating the
density of nodes from the phylogenetic root to the tip, while the
log-transformed equal splits (logES; also referred to as diversification
rate/DR) is derived from the sum of edge lengths from each tip towards
the root, with each edge towards the root having the length
down-weighted (Jetz et al. 2012; Quintero and Jetz 2018; Rabosky et al.
2018). Crucially, studies have suggested that DR and ND (henceforth
referred to as \(\lambda_{DR}\) and \(\lambda_{ND}\)) are more
reflective estimates of speciation than diversification. Because
\(\lambda_{DR}\) and \(\lambda_{ND}\) cannot account for whole-clade
extinctions and thus under-estimate extinction rate, which makes the
composite measure of diversification more dependent on speciation
(Belmaker and Jetz 2015; Title and Rabosky 2018). Therefore,
\(\lambda_{DR}\) is a measure of speciation rate more heavily weighted
to recent speciation events while \(\lambda_{ND}\) measures speciation
across the root-to-tip path. These tip-rate measures are alternatives to
state-dependent diversification models such as Quantitative State
Speciation-Extinction (QuaSSE); but, based on previous simulation
studies, \(\lambda_{DR}\) and \(\lambda_{ND}\) are robust and intuitive
measures that provide high power and low false discovery rate with large
phylogenies when incorporated into Phylogenetic Generalized Least
Squares (PGLS) models (Harvey Michael et al. 2017).

We used BAMM to model the dynamics of speciation and extinction across
the 101 phylogenetic trees (one MCC tree and 100 random draws of the
posterior). This software uses a Bayesian approach (reversible-jump
Markov Chain Monte Carlo) to generate probability distributions of
evolutionary rate-shift configurations with variable speciation and
extinction rates (Rabosky 2014). These models provide tip-rate estimates
of speciation and extinction rate that can be easily used in comparative
analyses. The parameters of the 100 BAMM runs are detailed in full in
the
\href{https://justincally.github.io/SexualSelection_Speciation/#set_up_bamm_parameters}{Supplementary Information};
briefly, we used a time-variable model with the prior expected number of
evolutionary rate shifts set at 100 and prior rates set from the initial
tip-level estimates of speciation and extinction using the
\texttt{BAMMtools} R package (Rabosky et al. 2014). BAMM models were run
independently for the 101 phylogenetic trees for 100 million
generations. Given the computationally intensive nature of BAMM, runs
were conducted across multiple CPUs. Important BAMM parameters
(log-likelihood and number of rate shifts) reached convergence with
effective sample size (ESS) of MCMC (Markov Chain Monte Carlo) samples
surpassing 200; an arbitrary value, above which posterior distributions
can often be accurately inferred
(\href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S3},
\href{https://justincally.github.io/SexualSelection_Speciation/#bamm_measures_of_speciation_and_extinction}{Table S4}).
Further details of BAMM parameters and output are available in the
\href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information},
with tip-rate means and variances provided. Additionally, given the
variability in BAMM estimates, we also provide analysis of BAMM shift
configurations and tip-rate estimates from our run on the MCC tree and
within a BAMM run on the MCC tree from a genetic-only phylogeny across
all birds (Harvey et al. 2017). All analyses were conducted on
log-rates. \phantomsection

\hypertarget{phylogenetic-comparative-analysis}{%
\subsection{Phylogenetic comparative
analysis}\label{phylogenetic-comparative-analysis}}

To test the association between speciation/extinction and sexual
selection, environmental variability and their interaction, we used
phylogenetic least squares (PGLS) models in the \texttt{nlme} package
(Pinheiro et al. 2018). Firstly, we conducted model selection to compare
models in which \(\lambda_{DR}\), \(\lambda_{ND}\), \(\lambda_{BAMM}\)
or \(\mu_{BAMM}\) were the response variable: these tip-rate estimates
all came from the same MCC tree (derived from 2,500 draws of the
posterior distribution (Jetz et al. 2012)). For models of
\(\lambda_{BAMM}\) and \(\mu_{BAMM}\), we used the inverse of the
variance associated with each tip rate estimate as weights, to account
for the variable precision of the estimates provided by BAMM. For each
response variable, we conducted model selection to compare models with
different combinations of predictor variables. The most complex model in
each set under comparison contained one of the measures of sexual
selection (sexual dichromatism or the index of male-biased sexual
selection), all of the environmental measures (i.e.~log-transformed
range size, seasonal temperature variation, spatial temperature
variation, long-term temperature variation, and NPP), and all of the
2-way interactions between sexual selection and each of the
environmental measures. The simpler models contained all of the same
main effects, but had fewer 2-way interaction terms (potentially none).
Model selection was done in \texttt{MuMIn} using the \texttt{dredge}
function (Bartoń 2017). Using the terms from the top-ranked model
(ranked by AICc), we ran the equivalent model for each of the 1,000
phylogenetic trees used to derive \(\lambda_{DR}\), \(\lambda_{ND}\) and
each of the 100 trees used to derive \(\lambda_{BAMM}\) and
\(\mu_{BAMM}\). Additionally, we investigated the effect of the
individual variables used to derive the index of male-biased sexual
selection on speciation rate. For these pgls models we replaced the
composite index score with the individual biological variable (sexual
size dimorphism, social polygyny and {[}lack of{]} paternal care) and
ran the equivalent model for 300 phylogenetic trees used to derive
\(\lambda_{DR}\), \(\lambda_{ND}\) and 100 trees used to derive
\(\lambda_{BAMM}\).

Across all our analyses we corrected for the phylogenetic signal. Our
models used the unique response variables and correlation structure for
a given phylogenetic tree. Specifically, for models using tip-rate
metrics (\(\lambda_{DR}\), \(\lambda_{ND}\)), we estimated the
phylogenetic signal independently for each of the 1,000 trees/models.
Phylogenetic signal was estimated as Pagel's \(\lambda\) (Pagel 1999)
using the \texttt{corPagel} function in the \texttt{ape} package
(Paradis et al. 2004). Alternatively, for models using speciation and
extinction estimates derived using BAMM (\(\lambda_{BAMM}\) and
\(\mu_{BAMM}\)), we found that \(\lambda\) was consistently estimated at
1 and hence assumed Brownian motion (using the \texttt{corBrownian}
function) to estimate the correlation structure. This method enabled us
to present model estimates for an MCC tree alongside 1,000/100 trees
from the posterior distribution of trees to account for phylogenetic
uncertainty. This approach was repeated on three datasets corresponding
to each measure of sexual selection: dichromatism derived from RGB
values of images (\emph{n} = 5,812); dichromatism from spectrophotometry
(\emph{n} = 581) and the index of male-biased sexual selection (\emph{n}
= 2,465).

Finally, using the subset of species with an index of male-biased sexual
selection, we conducted a phylogenetic path analysis using the
\texttt{phylopath\ R} package (Bijl 2018). The phylogenetic path
analysis was used to assess causal paths between variables unable to be
modelled within the univariate response of PGLS. That is, a phylogenetic
path analysis allowed us to model relationships between the predictor
variables used in our PGLS analysis as we anticipate environmental
variability, sexual dichromatism/selection, and range size to have
effects on each other and not just on speciation rate. To minimise path
complexity, we used temperature seasonality (BIO4) as the single measure
for environmental variability, \(\lambda_{DR}\) as the single measure of
speciation, and the tip-rates from the MCC tree. Further details of the
path analysis, including our rationale for each path's directions, can
be found within the
\href{https://justincally.github.io/SexualSelection_Speciation/}{Supplementary Information}
along with all other analyses and the relevant \texttt{R} code to
reproduce results. \phantomsection

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{male-biased-sexual-selection-but-not-sexual-dichromatism-affects-speciation}{%
\subsection{Male-biased sexual selection, but not sexual dichromatism,
affects
speciation}\label{male-biased-sexual-selection-but-not-sexual-dichromatism-affects-speciation}}

We examined the effect of sexual selection on speciation and extinction
rate in 97\% of passerines (\emph{n} = 5,812 species; 58\% of all birds;
\autoref{Speciation_Figure}). We found a significant positive
association between the index of male-biased sexual selection (\emph{n}
= 2,465) and \(\lambda_{DR}\) from the maximum clade credibility (MCC)
tree (\(\beta\) = \(3.89 \times 10^{-2}\), p = 0.01;
\autoref{model_results}b). However, this association was not significant
for the other two measures of speciation rate (\(\lambda_{ND}\):
\(\beta\) = \(4.38 \times 10^{-4}\), p = 0.35; \(\lambda_{BAMM}\):
\(\beta\) = \(9.42 \times 10^{-4}\), p = 0.76;
\autoref{model_results}b). When we took into account phylogenetic
uncertainty by running the models using 1,000 trees, the distribution of
estimates from PGLS models was similar to the estimate from the MCC
tree: among the 1,000 trees there was a positive association between
sexual selection and \(\lambda_{DR}\) (highest posterior density (HPD)
Interval = \(4.51 \times 10^{-3}\), \(5.72 \times 10^{-2}\)), and the
distribution skewed towards a positive association between sexual
selection and \(\lambda_{ND}\) (HPD Interval = \(-5.04 \times 10^{-4}\),
\(1.58 \times 10^{-3}\)) as well as the 100 models using
\(\lambda_{BAMM}\) (HPD Interval = \(-1.30 \times 10^{-2}\),
\(3.09 \times 10^{-2}\);
\href{https://justincally.github.io/SexualSelection_Speciation/#analysis_using_male-biased_measure_of_sexual_selection}{Table S15}).

We investigated which of the three variables comprising the index of
male-biased sexual selection was driving the association observed with
\(\lambda_{DR}\). Our results over 300 trees showed that this pattern is
mainly driven by the sexual size dimorphism component (HPD Interval =
\(8.53 \times 10^{-1}\), 3.11), with the effects of other components
overlapping zero; paternal care (HPD Interval =
\(-1.78 \times 10^{-1}\), \(7.90 \times 10^{-3}\)) and mating system
(HPD Interval= \(-7.35 \times 10^{-2}\), \(4.32 \times 10^{-2}\)).
Importantly, the association between sexual size dimorphism and
speciation rates is also present when using \(\lambda_{ND}\) (HPD
Interval = \(1.80 \times 10^{-1}\), \(6.38 \times 10^{-1}\)), but not
when using \(\lambda_{BAMM}\) (HPD Interval = -1.49,
\(7.45 \times 10^{-1}\), \autoref{SS_components}).

>>>>

\usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath} \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}

\pretitle{\begin{flushleft}\LARGE}
\posttitle{\par\end{flushleft}\vskip 0.5em}
\preauthor{\begin{flushleft}\large \lineskip 0.5em}
\postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
\postdate{\par\end{flushleft}}

\usepackage{fancyhdr}

\pagestyle{fancy} \fancyhead[L]{\textsl{\leftmark}}
\rhead[]{Sexual selection predicts speciation}

\usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{setspace}

\onehalfspacing

\usepackage[font={small}]{caption} \usepackage{epigraph}

\setlength\epigraphwidth{5.25in} \setlength\epigraphrule{0pt}
\setlength\beforeepigraphskip{1\baselineskip}
\setlength\afterepigraphskip{2\baselineskip}

\renewcommand*{\textflush}{flushright} \renewcommand*{\epigraphsize}{\footnotesize\itshape} \usepackage{titlesec} \titleformat{\section}{\normalfont\scshape\LARGE}{\thesection}{1em}{} \titleformat{\subsection}{\normalfont\Large}{\thesubsection}{1em}{} \titleformat{\subsubsection}{\normalfont\large}{\thesubsubsection}{1em}{} \usepackage{abstract} \renewcommand{\abstractnamefont}{\normalfont\Large} \usepackage[hang, flushmargin]{footmisc} \usepackage{hyperref}

\hypersetup{colorlinks=true, hyperfootnotes=false, linkcolor=blue, filecolor=blue, urlcolor=blue}

\usepackage{xcolor} \renewcommand\thefootnote{\textcolor{black}{\arabic{footnote}}}

\raggedbottom

\begin{document}
\maketitle

@@
